Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees

### 2023-02-21 script/manuscript_MEE.Rmd
final touches for the night

@@

bee_obs <- read.csv(
  file.path('../data', files[grep('queen_observations', files)]) ) %>% 
  filter(caste == 'q') %>% 
  mutate(species = gsub('\\..*$', '', species)) %>% 
  select(site, doy, week, tot.obs.length, species, plant.species,
         resource.coll:pollen.collected, fl.switch.1:fl.switch.2.resource.coll) %>% 
  mutate(across(resource.coll:fl.switch.2.resource.coll, ~ na_if(.x, ''))) %>% 
  pivot_longer(c(plant.species, fl.switch.1, fl.switch.2), names_to = 'observation', values_to = 'plant.species') %>% 
  drop_na(plant.species) %>% 
  group_by(site, week, species, plant.species) %>% 
  count(name = 'interactions') %>% 
  left_join(., obs_time, by = c("site", "week")) %>% 
  mutate(period = case_when(
    week %in% 3:5 ~ 'Early',
    week %in% 6:8 ~ 'Mid',
    week %in% 9:11 ~ 'Late'
  ))

@@

To ascertain the extent to which records of multiple species in a family, which were suspected to be sampling artefacts occurred in molecular samples an index of similarity, ala jaccard, the affinity index was used to assess co-occurrence (@mainali2022better, @mainali2022alpha). Numerous taxa from the family Ranunculaceae Jussieu (*Caltha* L. sp., *Thalictrum* L. spp., *Trollius* L. sp., *Aquilegia* L. spp.), had $\alpha$ scores which indicated that they are only present when a more common confamilial taxa *Delphinium barbeyi* (Huth) Huth *nuttallianum* Pritz. were recorded. A similar relationship was observed in the Hydrophyllaceae R.Br. with samples placed in *Nemophila* Nutt., which only occurred when the more abundant *Hydrophyllum* L. species were present. The size of flower of *Nemophila breviflora* A. Gray make it unlikely to be visited by Bumble Bees, and it is a false positive. The floral morphology of *Thalictrum*  spp. also makes it unlikely to be visited (BELEIVE JANE SAID THIS- IF NOT STRIKE OUT), and while evidence of visits to Caltha and Trollius are lacking, due to the association between the reads these results appear unlikely.

@@

```{r Build reference dataset for reclassification process}

# write.csv(blst, '../data/blst4experts.csv' , row.names = F)

pollen_sample_info <- readxl::read_xlsx(file.path('../data', 
  list.files(path = '../data', pattern = 'xlsx') )) %>% 
  select(site, date, sample.id) %>% 
  distinct() %>% 
  filter(!date %in% c('42170', '42169'))

files <- list.files(path = '../data')

flr_date_ranges <- read.csv(
  file.path('../data',  files[grep('flower_ranks', files)] )) %>% 
  select(site, doy, week, plant.species:abun.rank) %>% 
  group_by(plant.species, site) %>% 
  select(site, week, doy, plant.species) %>% 
  mutate(Genus = str_extract(plant.species, '^.+?(?=[.]|$)')) %>% 
  mutate(Genus = case_when(
    Genus == 'Erythrocoma' ~ 'Geum', 
    Genus == 'Amerosedum' ~ 'Sedum', 
    Genus == 'Pentaphylloides' ~ 'Dasiphora',
    TRUE ~ Genus
    )) %>% 
  mutate(plant.species = case_when(
    plant.species == 'Lathyrus.leucanthus' ~ 'Lathyrus.lanszwertii',
    plant.species == 'Lupinus.bakeri' ~ 'Lupinus.sericeus',
    TRUE ~ plant.species
  )) %>% distinct()
 # this provides us with a data.frame of the ACTUAL flowering dates!

week_tab <- flr_date_ranges %>% 
  ungroup() %>% 
  select(doy, week) %>% 
  distinct()

flr_date_ranges <- select(flr_date_ranges, -doy)

blst_expert <- blst %>% 
  select(species, sample.id, TAXON_NEW, Genus, Family, Prcnt_seqs, Prcnt_Seqs_Class) %>% 
  mutate(RecordID = 1:n()) %>% 
  left_join(., pollen_sample_info,  by = 'sample.id') %>% 
  relocate(site:date, .before = TAXON_NEW) %>% 
  mutate(doy = lubridate::yday(date)) %>% 
  left_join(., week_tab, by = 'doy') %>% 
  left_join(., flr_date_ranges, by = c('site', 'Genus', 'week')) %>% 
  relocate(sample.id, .before = plant.species)

rm(corb, week_tab, flr_date_ranges, pollen_sample_info)

  # reduce to temporally similar results

######## SAMPLE 55 SHOULD BE SOMEWHERE HERE. 
duplicate_reps_in_sample <- blst_expert %>%

  # identify the duplicate samples
  drop_na(plant.species) %>% 
  group_by(RecordID) %>% 
  distinct(plant.species, .keep_all = T) %>% 
  group_by(sample.id, TAXON_NEW) %>% 
  filter(n() >= 2)

dupe_ids <- duplicate_reps_in_sample %>% pull(RecordID) %>% unique()
dupe_samples <- duplicate_reps_in_sample %>% pull(sample.id) %>% unique()

  # now split the reads assigned to them into each constituent taxon ~ 'unwise King Solomon'
duplicate_reps_in_sample <- duplicate_reps_in_sample %>% 
  ungroup() %>% 
  group_by(sample.id, Genus) %>% 
  mutate(Prcnt_seqs_Reclass = 
           ifelse(n() == 2, Prcnt_seqs/2, sum(Prcnt_seqs)/n_distinct(plant.species)/2 ),
         TAXON_EXPERT = plant.species) %>% 
  distinct(sample.id, plant.species, .keep_all = T)

ranunc_splash <- c('Caltha', 'Trollius', 'Thalictrum', 'Aquilegia')

blst_reclass_expert <- blst_expert %>% 
  filter(!RecordID %in% dupe_ids) %>% 
  distinct() %>% 
  rowwise() %>% 
  mutate(Genus = str_trim(Genus)) %>%

  mutate(TAXON_EXPERT = case_when(

    sample.id == 3 & Family == 'Asteraceae' ~  'Taraxacum.officinale',
    sample.id == 3 & Family %in% c('Rosaceae', 'Onagraceae') ~  'Potentilla.pulcherrima',
    sample.id == 3 & Genus == 'Delphinium' ~ plant.species,       ## COMPLETE ##
    sample.id == 4 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 4 ~ plant.species,                               ## COMPLETE ##
    sample.id == 6 ~ plant.species,                               ## COMPLETE ##
    sample.id == 9 & TAXON_NEW == 'Symphyotrichum foliaceum' ~ 'Erigeron speciosus',
    sample.id == 9 ~ plant.species,                               ## COMPLETE ##

    sample.id == 10 & Family == 'Asteraceae' ~  'Taraxacum.officinale',
  #  sample.id == 10 & Family == 'Onagraceae' ~ ''
    sample.id == 10 ~ plant.species, # has many late season taxa

    sample.id == 14 & Genus == 'Symphyotrichum' ~ 'Senecio.integerrimus',
    sample.id == 14 ~ plant.species,                              ## COMPLETE ##
    sample.id == 15 & Family %in% c('Onagraceae', 'Rosaceae')  ~ 'Potential pulcherrima',
    sample.id == 15 & Family == 'Asteraceae' ~ 'Senecio.integerrimus',
    sample.id == 15 ~ plant.species,                              ## COMPLETE ## 
    sample.id == 18 & Family == 'Ericaceae' ~ 'Ericaceae', # ONLY TO FAMILY
    sample.id == 18 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', 
    sample.id == 18 & Family == 'Asteraceae' ~ 'Senecio.integerrimus',
    sample.id == 18 ~ plant.species,                              ## COMPLETE ##

    sample.id == 19 & Family == 'Onagraceae' ~ 'Epilobium',
    sample.id == 19 & Genus == 'Pseudognaphalium' ~ 'Senecio.integerrimus',
    sample.id == 19 & Genus == 'Nemophila' ~ 'Hydrophyllum.fendleri',
    sample.id == 19 ~ plant.species,                              ## COMPLETE ##
    sample.id == 20 & TAXON_NEW == 'Hydrophyllum canadense' ~ 'Hydrophyllum fendleri',
    sample.id == 20 & Family == 'Asteraceae' ~ 'Senecio.integerrimus',
    sample.id == 20 ~ plant.species,                              ## COMPLETE ##
    sample.id == 21 & Family == 'Asteraceae' ~ 'Senecio.integerrimus',
    sample.id == 21 ~ plant.species,                              ## COMPLETE ##
    sample.id == 23 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 23 ~ plant.species,                              ## COMPLETE ##
    sample.id == 24 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 24 ~ plant.species,                              ## COMPLETE ##
    sample.id == 25 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 25 ~ plant.species,                              ## COMPELTE ##
    sample.id == 26 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 26 ~ plant.species,                              ## COMPLETE ##
    sample.id == 27 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 27 ~ plant.species,                              ## COMPLETE ##
    sample.id == 28 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 28 ~ plant.species,                              ## COMPLETE ##
    sample.id == 29 ~ plant.species,                              ## COMPLETE ##
    sample.id == 30 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', 
    sample.id == 30 ~ plant.species,                              ## COMPLETE ##
    sample.id == 30 & Genus == 'Erigeron' ~ 'Erigeron elatior',
    sample.id == 32 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 32 ~ plant.species,                              ## COMPLETE ##

    sample.id == 33 ~ plant.species, # Symphyotrichum foliaceum check on day 166

    sample.id == 35 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 35 ~ plant.species,                              ## COMPLETE ##
    sample.id == 36 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis',  
    sample.id == 36 ~ plant.species,                              ## COMPLETE ##
    sample.id == 39 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 39 ~ plant.species,                              ## COMPLETE ##
    sample.id == 40 ~ plant.species,                              ## COMPLETE ##
    sample.id == 41 & Genus == 'Cynoglossum' ~ 'Mertensia.ciliata', 
    sample.id == 41 & Genus == 'Viola' ~ 'Viola adunca',          ## COMPLETE ##
    sample.id == 41 ~ plant.species,                              ## COMPLETE ##
    sample.id == 42 ~ plant.species,
    sample.id == 43 ~ plant.species,                              ## COMPLETE ##
    sample.id == 44 ~ plant.species,
    sample.id == 45 & Family == 'Asteraceae' ~ 'Senecio.integerrimus',
    sample.id == 45 ~ plant.species,                              ## COMPLETE ##
    sample.id == 46 & Family == 'Asteraceae' ~ 'Senecio.integerrimus',
    sample.id == 46 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', 
    sample.id == 46 ~ plant.species,                              ## COMPLETE ##
    sample.id == 49 & Genus %in% ranunc_splash ~ 'Delphinium.barbeyi',
    sample.id == 49 & Genus == 'Viola' ~ 'Viola.adunca',
    sample.id == 49 ~ plant.species,                              ## COMPLETE ##

    sample.id == 51 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 51 ~ plant.species,  ## Helianthella quinquenervis @ DOY 166 ##
    sample.id == 52 ~ plant.species,    ## Erigeron elatior  on DOY 163 @ AVERY WASH ##

    sample.id == 54 ~ plant.species,                              ## COMPLETE ##
    sample.id == 55 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', 
    sample.id == 55 ~ plant.species,                              ## COMPLETE ##
    sample.id == 56 & Genus == 'Nemophila' ~ 'Hydrophyllum fendleri',
    sample.id == 56 & Family == 'Asteraceae' ~ 'Senecio.integerrimus',
    sample.id == 56 ~ plant.species,                              ## COMPLETE ##
    sample.id == 60 & Genus == 'Viola' ~ 'Viola.adunca',          ## COMPLETE ##
    sample.id == 60 ~ plant.species,
    sample.id == 65 & Genus == 'Viola' ~ 'Viola.praemorsa',       ## COMPLETE ##
    sample.id == 65 ~ plant.species,

   TRUE ~ as.character(TAXON_NEW)
  ))  %>% 
  mutate( 
    TAXON_EXPERT = str_replace_all(TAXON_EXPERT, '[.]', ' '),
    TAXON_EXPERT = if_else(is.na(TAXON_EXPERT), TAXON_NEW, TAXON_EXPERT),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Holodiscus argenteus', 'Potentilla pulcherrima'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Scabrethia scabra', 'Wyethia arizonica'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Geum ternatum', 'Geum triflorum'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Erigeron grandiflorus', 'Erigeron elatior'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Agastache pallidiflora', 'Pedicularis'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Lithophragma parviflorum', 'Lithophragma glabrum'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 
                               'Arenaria globiflora|Minuartia recurva|Sagina procumbens', 'Eremogone congesta'),
    TAXON_EXPERT = if_else(doy <= 171 & Genus == 'Cynoglossum', 
                           'Mertensia fusiformis', TAXON_EXPERT), 
    TAXON_EXPERT = if_else(doy < 190 & TAXON_EXPERT == 'Lupinus argenteus', 
                           'Lupinus sericeus', TAXON_EXPERT))  %>%

  # use this to sum up the total sequences by original read record
  distinct(sample.id, TAXON_NEW, .keep_all = T) %>% 
  group_by(sample.id, TAXON_EXPERT) %>% 
  mutate(Prcnt_seqs_Reclass = sum(Prcnt_seqs)) %>%

  # now reduce to the amount of rows of the proper classification
  ungroup() %>% 
  distinct(sample.id, TAXON_EXPERT, .keep_all = T) %>% # what is going on with sample 56 then?

  # add back on the multiple groups
  bind_rows(., duplicate_reps_in_sample)  %>% 
  arrange(sample.id, -Prcnt_seqs_Reclass)

rm(ranunc_splash)

# now we split up the reads assigned to conflicted groups  # e.g. Cynoglossum amplifolium in sample 65
b <- blst_reclass_expert %>% 
  filter(sample.id %in% dupe_samples) %>% 
  group_by(sample.id, Family) %>% 
  filter(n() >= 3) %>% 
  mutate(plant.species = na_if(plant.species, "Thalictrum.fendleri")) %>% 
  mutate(Seqs2distribute = sum(Prcnt_seqs_Reclass))

records2drop <- b %>% pull(RecordID)

b <- b %>% 
  drop_na(plant.species) %>% 
  mutate(Prcnt_seqs_Reclass = Seqs2distribute/n_distinct(TAXON_EXPERT)) %>% 
  select(-Seqs2distribute)

blst_reclass_expert <- blst_reclass_expert %>% 
  filter(!RecordID %in% records2drop) %>% 
  bind_rows(., b) %>% 
  arrange(sample.id, -Prcnt_seqs_Reclass)

rm(records2drop, b)

# 55 SHOULD NOT BE IN HERE, HOW IS NOT CAUGHT IN THE ABOVE EXCEPTION???? #
# seems i may not have either delphinium present in that week? and  if falls between two. 
# deal with sample 55 one off, split the delphinium gracile

s55 <- blst_reclass_expert %>% 
  filter(sample.id == 55) %>% 
  group_by(Genus) %>% 
  mutate(Seqs2distribute = sum(Prcnt_seqs_Reclass)) %>% 
  filter(TAXON_EXPERT != 'Delphinium gracile') %>% 
  mutate(Prcnt_seqs_Reclass = Seqs2distribute/n_distinct(TAXON_EXPERT)) %>% 
  select(-Seqs2distribute)

blst_reclass_expert <- blst_reclass_expert %>% 
  filter(sample.id != 55) %>% 
  bind_rows(., s55)  %>% 
  arrange(sample.id, -Prcnt_seqs_Reclass)

rm(s55)

# we will split the reads for the Cynoglossum detected in sample 3.

blst_reclass_expert <- blst_reclass_expert %>% 
  filter(sample.id == 3 & TAXON_EXPERT == 'Cynoglossum pringlei') %>% 
  slice(rep(1:n(), each=2)) %>% 
  mutate(Prcnt_seqs_Reclass = Prcnt_seqs_Reclass/2,
         dummyID = 1:n(),
         TAXON_EXPERT = if_else(dummyID == 1, 'Mertensia fusiformis', 'Mertensia cilita')) %>% 
  select(-dummyID) %>% 
  bind_rows(., filter(
              blst_reclass_expert, sample.id == 3 & TAXON_EXPERT != 'Cynoglossum pringlei'),
            filter(blst_reclass_expert, sample.id != 3))

# re class the totals percent of all sequences to sum to 100%

blst_reclass_expert <- blst_reclass_expert %>% 
  group_by(sample.id) %>% 
  mutate(multiplier = 100/ sum(Prcnt_seqs_Reclass),
         Prcnt_seqs_Reclass2 = Prcnt_seqs_Reclass * multiplier)

yargh <- split(blst_reclass_expert, f = blst_reclass_expert$sample.id)

rm(pollen_sample_info)
```

@@

```{r}
a <- read.csv('../data/herbaria_4_acknowledgements.csv') %>% 
  pull() 
```

>>>>

### 2.7 | Integrated Observational, Molecular, and Palynological Corbiculae

In order to precisely classify the contents of each corbiculae load the sequences classified by molecular methods were compared with the fieldwork which at a very fine resolution, recorded the presence and absence of species and their duration of flowering, and was interpreted ala the computer derived temporal and spatial data sets. The quantitative counts of grains from microscopy, were combined with the semi-quantitative sequencing results, to estimate the abundance of each identified species in each corbiculae load.

To reclassify the sequence reads, these data were combined with the flora observation data, and mapped by genus. If more than one species in the genus was flowering at that time and site, than the reads were split evenly between the taxa. For sequence data which did not match at the genus level, a user subjectively scored them based on the species composition and phenological activity at each plot, the Queen interaction data, and pollen assignments. To estimate the abundance of each of these species in the corbiculae loads, these data were combined with the microscopy data. For each morphotype detected in pollen, and each classified sequence read which was not detected via microscopy, they were given a value of 0.5% to indicate there trace presences. When more than a single species belonged to a morphotype group in a single sample, the quantitative values from the morphological work were multiplied by the relative sequence abundance of each species in the load. All final compositions were standardized to a sum of 100%, by adding or subtracting the differences (induced by classifying records as 'trace') to all species with abundances > 1%.

@@

bee_obs <- read.csv(
  file.path('../data', files[grep('queen_observations', files)]) ) %>% 
  filter(caste == 'q') %>% 
  mutate(species = gsub('\\..*$', '', species)) %>% 
  select(site, doy, week, tot.obs.length, species, plant.species,
         resource.coll:pollen.collected, fl.switch.1:fl.switch.2.resource.coll) %>% 
  mutate(across(where(is.character), ~ na_if(., ''))) %>% 
  pivot_longer(c(plant.species, fl.switch.1, fl.switch.2), names_to = 'observation', values_to = 'plant.species') %>% 
  drop_na(plant.species) %>% 
  group_by(site, week, species, plant.species) %>% 
  count(name = 'interactions') %>% 
  left_join(., obs_time, by = c("site", "week")) %>% 
  mutate(period = case_when(
    week %in% 3:5 ~ 'Early',
    week %in% 6:8 ~ 'Mid',
    week %in% 9:11 ~ 'Late'
  ))

@@

To ascertain the extent to which records of multiple species in a family, which were suspected to be sampling artefacts occurred in molecular samples an index of similarity, ala jaccard, the affinity index was used to assess co-occurrence (@mainali2022better, @mainali2022alpha). Numerous taxa from the family Ranunculaceae Jussieu (*Caltha* L. sp., *Thalictrum* L. spp., *Trollius* L. sp., *Aquilegia* L. spp.), had $\alpha$ scores which indicated that they are only present when a more common confamilial taxa *Delphinium barbeyi* (Huth) Huth *nuttallianum* Pritz. were recorded. A similar relationship was observed in the Hydrophyllaceae R.Br. with samples placed in *Nemophila* Nutt., which only occurred when the more abundant *Hydrophyllum* L. species were present. The size of flower of *Nemophila breviflora* A. Gray make it unlikely to be visited by Bumble Bees, and it is a false positive. The floral morphology of *Thalictrum*  spp. also makes it unlikely to be visited (BELEIVE JANE SAID THIS- IF NOT STRIKE OUT), and while evidence of visits to *Caltha* and *Trollius* are lacking, due to the association between the reads these results appear unlikely.

@@

```{r Build reference dataset for reclassification process}
reclassified <- read.csv('../data/Fully_Integrated_Corbiculae.csv')
```

@@

```{r}
a <- pull(read.csv('../data/herbaria_4_acknowledgements.csv'))
```
